---
title: "Getting started with monitoraSom"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with monitoraSom}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  out.width = "100%",
  dpi = 96
)
```

```{r library, message = FALSE}
library(monitoraSom)
```

In this document, we, the monitoraSom developers, will walk you through a
complete template matching workflow. You will extract templates from focal
recordings, search soundscape recordings for similar signals, and inspect the
detections that come out. Most steps are shown as code you can read; the steps that need no audio run
for real, on the data bundled with the package. To follow the full example
with its recordings, download the example project first (see the download
section below).

## What this package does

`monitoraSom` turns raw soundscape recordings into ecological data through
template matching. A template is a short, curated recording of the sound you
are looking for. The engine slides that template over every recording in a
corpus and scores the acoustic similarity at each position. Peaks in that score
become candidate detections. Detections are then validated against manually
segmented regions of interest (ROIs), and the validated set is scored with
standard detection metrics.

The workflow runs in nine steps.

1. `set_workspace()` creates the project directory tree and its DuckDB stores.
2. `fetch_soundscape_metadata()` reads the recordings and builds the corpus table.
3. `launch_segmentation_app()` opens the interface where you draw ROIs by hand.
4. `fetch_rois()` reads those ROIs back into R.
5. `export_templates()` cuts the ROIs you marked as templates into WAV files.
6. `fetch_template_metadata()` reads the template library.
7. `fetch_match_grid()` pairs every soundscape with every template.
8. `run_matching()` scores each pair, and `fetch_score_peaks()` turns the score
   series into detections.
9. `validate_by_overlap()` and `diagnostic_validations()` measure how well the
   templates performed.

## What this dataset is

The worked example uses the original data from the monitoraSom 1.0.0
publication: recordings of *Basileuterus culicivorus* (the golden-crowned
warbler). It pairs two kinds of material. Focal recordings, taken by observers
in the field with handheld recorders, from which the templates are obtained, and
soundscape recordings, taken by autonomous recorders in passive acoustic
monitoring, where we search for occurrences of the signals represented by the
templates.

|                      |                                                 |
| -------------------- | ----------------------------------------------- |
| Focal recordings     | 2 (`Bcu_1`, `Bcu_2`), the source of the templates |
| Search soundscapes   | 12, where we look for similar songs             |
| ROIs                 | 71 (18 focal + 53 search)                       |
| Templates            | 6 (the Substructure C ROIs)                    |
| Species              | ***B. culicivorus***                            |

> **Important:** only *B. culicivorus* songs were segmented. Precision and
> recall are meaningful for the target species' song only; calls of the same
> species and vocalizations of other species are not represented in this
> example. One recording carries the mark `no signals of interest`, meaning
> it was reviewed and no song of the target species was found.

## Download the example project

To follow the example with its recordings, get the example project with
`fetch_example_data()`. It downloads the project into your user cache folder
and returns its path, ready for `setwd()` or the `project_path` argument of
the pipeline functions. `is_example_data_cached()` reports whether the
project is already there. The download is opt-in: nothing in the package
triggers it on its own.

```{r download-example, eval = FALSE}
is_example_data_cached()
example_project <- fetch_example_data()
setwd(example_project)
```

The chunk above does not run while this vignette is built, so the build stays
network-free. Run it once, by hand, in your own session. If the host is
unreachable, the download times out, or the bundle is malformed,
`fetch_example_data()` prints an informative message and returns `NULL`; it
never errors. The bundle lands at
`<your cache>/basileuterus-culicivorus` and stays there across R sessions,
so later calls are a no-op unless you pass `overwrite = TRUE`. To use a local
copy or a mirror instead of the public release, set
`options(monitoraSom.data_url = "file:///path/to/bundle.tar.gz")` first.

A machine with no network can still follow this vignette: the audio is not
needed for the bundled-table sections, which run for real from "The project
directory" onwards.

## The project directory

The downloaded project is a small real project: the 14 recordings, a ready
`rois.duckdb` and the walkthrough in three formats. The audio is far too
large to ship inside a package, so you get it as a download.

```
bcu/                     project root
├── bcu.qmd              the walkthrough document
├── recordings/          2 focal recordings, the template source
│   ├── Bcu_1.wav
│   └── Bcu_2.wav
├── soundscapes/         12 soundscape recordings to search
│   └── W54393S25597_…wav
└── rois.duckdb          71 manually segmented ROIs
```

The paths inside `rois.duckdb` are relative to the directory that contains it.
You can move the folder anywhere and everything keeps working. This matters
when you collaborate with others. Downloading the project is covered in the
download section above.

```{r bundled-data}
data(df_soundscapes)
data(df_templates)
data(df_rois)
data(df_detecs)

c(
  soundscapes = nrow(df_soundscapes),
  templates = nrow(df_templates),
  ground_truth_rois = nrow(df_rois),
  detections = nrow(df_detecs)
)
```

The package itself carries a small, complete result set from that same study.
The audio is not included, but every table the audio produced is: 12
soundscapes searched with 6 templates, `r nrow(df_detecs)` candidate
detections, and `r nrow(df_rois)` ground truth ROIs. The chunks from here to
the end of the diagnostics run on these bundled tables, so you can execute the
vignette end to end without downloading anything.

## Preparation

The first step prepares the environment so the active R session knows where
the files to read are and where the exported files go. In the downloaded
example, the walkthrough document sits at the project root, so the relative
paths resolve from there.

```{r workspace, eval = FALSE}
# Create the project tree and its DuckDB stores:
set_workspace("my_project")
```

## Reading the inputs

With the environment ready, we read the inputs of the workflow: the metadata
of the 12 soundscapes and the ROI table that the example ships with. The
`fetch_soundscape_metadata()` function walks the soundscapes folder, reads each
recording, and stores the metadata in a DuckDB database as well as returning
it to your R session.

This step matters because the metadata feed every later step, and a reading
problem that slips through here propagates to the rest of the analysis. If a
corrupted recording were silently skipped, the search would cover less audio
than you planned, and you would only notice through detections that never
appeared. So the function logs a recording that will not open, a truncated
metadata header, or a zero-byte file into an error log, without interrupting
the sweep of the remaining files. An empty folder, with no readable recording
at all, is an immediate error rather than an empty table that would only
denounce the problem much later.

```{r inputs, eval = FALSE}
df_soundscapes <- fetch_soundscape_metadata(
  soundscapes_path = "my_project/soundscapes",
  output_file = "my_project/soundscapes_metadata.duckdb"
)
```

The same care applies to the ROI table. When the path points to a folder with
several ROI stores, an unreadable file is not ignored, so no segmentation is
lost without notice. Types and variable compatibilities are checked, and
values that do not match the monitoraSom standard are converted to `NA` with a
warning that names the source file. The checks also read tables produced by
older package versions in the current format, and standardize recording paths,
so no ROI is orphaned by cosmetic differences in how a path is written.

```{r inputs-rois, eval = FALSE}
df_rois <- fetch_rois("my_project/rois.duckdb")
```

## The whole analysis in one command

`template_matching()` is the main function of the workflow. It reads the
recording and template metadata, builds the search grid, runs the matching,
and returns the detections, all in a single call.

```{r canonical-run, eval = FALSE}
df_detecs <- template_matching(
  soundscapes_path = "my_project/soundscapes/",
  templates_path = "my_project/templates/templates.duckdb",
  score_method = "fft", min_score = 0.4
)
```

The `score_method` argument chooses the search engine. `"cor"` computes the
Pearson correlation between the template and every sliding window of the
soundscape; it is the original reference implementation. `"fft"` computes the
same correlation through Fourier transforms. Both give the same result, to
differences around 1e-6, but `"fft"` is about ten times faster, which is why
it is the current default.

The call above leaves the results in your R session only. To save them to
disk, use the `output_db` argument.

```{r persisting, eval = FALSE}
template_matching(
  soundscapes_path = "my_project/soundscapes/",
  templates_path = "my_project/templates/templates.duckdb",
  min_score = 0.4,
  output_db = "my_project/detections.duckdb"
)
```

## The steps inside the orchestrator

The single call is the path we recommend, but it only runs, in sequence, the
four stages that make up the orchestrator. The chunk below shows those four
stages, plus the preparation that feeds them. The orchestrator covers
everything from `fetch_template_metadata()` down to `fetch_score_peaks()`;
a comment marks where it ends. The preparation before it and the validation
and diagnostics sections after it are separate steps that you call yourself.
Call the stages separately when you need to inspect or reuse an intermediate
result.

```{r expanded, eval = FALSE}
# The same workflow as the orchestrator, step by step:
set_workspace("my_project")
df_soundscapes <- fetch_soundscape_metadata("my_project/soundscapes/")

# Draw ROIs by hand; the app writes them straight to the ROI store.
launch_segmentation_app()

df_rois <- fetch_rois("my_project/rois.duckdb")

# The six templates are the Substructure C ROIs (the final, wider part):
df_template_rois <- subset(df_rois, roi_comment == "Substructure C")
export_templates(
  df_template_rois,
  templates_path = "my_project/templates/",
  create_dir = TRUE
)
df_templates <- fetch_template_metadata("my_project/templates/templates.duckdb")

# The four stages below are what template_matching() runs inside itself:
df_grid   <- fetch_match_grid(df_soundscapes, df_templates)   # 72 pairs
df_scores <- run_matching(df_grid, score_method = "fft")
df_detecs <- fetch_score_peaks(df_scores, min_score = 0.4)
# The orchestrator stops here. The validation and diagnostics sections
# below run after it, on the detections it returned.
```

Two things about the template step deserve your attention. First, the template
identity stays tied to its source recording: the template database records the
file, the instant and the frequency band of the cut, together with the hash of
the source audio. This lets you check that a template matches the segment you
drew; a loose audio file carries no such record of origin. Second, the cut is
derived deterministically from the ROI coordinates over the original audio:
every run produces exactly the same file, and re-exporting creates no
divergent versions.

Most workflows prefer `template_matching()` when:

- every soundscape in `df_soundscapes` is searched with every template in
  `df_templates`, which necessarily uses the full grid in `df_grid`; and
- the complete score series in `df_scores`, which can be heavy, is not needed,
  so retaining only the detections as the final product is enough.

Before launching a long run, two helpers are worth knowing.
`estimate_matching()` reports how big a search plan is, reading only the grid
and never the audio, so a surprising pair count surfaces before anything
expensive starts. `summarise_matching()` reports what a finished run produced.

```{r estimate}
data(df_grid)
estimate_matching(df_grid)
```

It returns the same figures invisibly as a one row `data.frame`, so
`quiet = TRUE` suits a script that wants the numbers without the report.

## Validation

Detections are candidates, not results: the matching measures similarity, and
similarity is not identification. To know which detections are real, you need
reference ROIs covering the same recordings the search went through.
`validate_by_overlap()` makes that judgment. A detection that overlaps an ROI
of the expected label is a true positive. One that overlaps nothing is a false
positive. An ROI that no detection reached is a false negative, which is the
part of the truth set the templates missed.

One rule decides which recordings take part: a recording counts only when its
review state is known. It either carries ROIs, or you marked it "no signals
of interest" in the segmentation app (`launch_segmentation_app()`) after a
full review. A recording with neither was never reviewed, so its detections
are excluded with a warning and never count as absence of the species. This
keeps "not validated yet" separate from "validated as absent".

```{r validate}
val <- validate_by_overlap(df_detecs, df_rois, validation_user = "vignette")

df_validated <- val$detections_validated
table(df_validated$validation)
```

False negatives are reported per template, because each template searched the
whole corpus on its own.

```{r false-negatives}
fn <- val$false_negatives

nrow(fn)
table(substr(fn$template_name, 1, 11))
```

The bundled detections were produced with a deliberately permissive threshold,
so false positives outnumber true positives by about six to one, and each
template still missed 18 or 19 ground truth ROIs. Raising the threshold trades
false positives for more misses. Choosing where to sit on that curve is what
the diagnostics are for.

## Diagnostics

`diagnostic_validations()` sweeps every observed score as a candidate threshold
and computes precision, recall and F1 at each one. It returns one diagnostic
per template, because templates are not interchangeable. Two cuts of the same
species can need very different thresholds.

```{r diagnostics}
diags <- diagnostic_validations(df_validated, build_plots = TRUE)

length(diags)
```

Each element carries the sweep itself, the score cut, the fitted logistic model
and a set of plots.

```{r diagnostics-structure}
names(diags[[1]])
```

The score cut is derived from the fitted logistic model, as the score at which
the model gives a detection a posterior probability of `pos_prob` of being a
true positive. That default is 0.95, which is a demanding requirement. Template
names encode the source recording, the time and frequency bounds of the cut
and the spectrogram parameters, so they are abbreviated below.

```{r auto-cut}
short_name <- function(x) substr(x, 1, 11)

data.frame(
  template = short_name(names(diags)),
  auto_cut = round(sapply(diags, `[[`, "score_cut"), 3),
  max_observed = round(
    sapply(diags, function(d) max(d$diagnostics$peak_score)), 3
  ),
  row.names = NULL
)
```

On this demo corpus the automatic cut sits above the highest score any template
actually reached. No detection can clear it, so no operating point is flagged
and the `selected` column of the sweep stays `FALSE` throughout. This is not a
failure of the fit. It is the model reporting that 95% confidence is not
attainable from six templates on twelve recordings.

Lowering `pos_prob` asks for less confidence and brings the cut back inside
the observed range.

```{r relaxed-cut}
diags_relaxed <- diagnostic_validations(
  df_validated,
  pos_prob = 0.5,
  build_plots = FALSE
)

summary_rows <- lapply(diags_relaxed, function(d) {
  best <- d$diagnostics[d$diagnostics$selected, ]
  if (nrow(best) == 0) {
    return(NULL)
  }
  best <- best[1, ]
  data.frame(
    template = short_name(best$template_name),
    score_cut = round(d$score_cut, 3),
    tp = best$tp,
    fp = best$fp,
    fn = best$fn,
    precision = round(best$precision, 3),
    recall = round(best$recall, 3),
    F1 = round(best$F1_score, 3)
  )
})

do.call(rbind, unname(summary_rows))
```

Five of the six templates now have an operating point, and every one of them
sits at perfect precision with low recall. That is the conservative end of the
curve, where a detection is almost certainly real but most of the
vocalizations are missed. Whether that suits your study is an ecological
question rather than a statistical one.

Every diagnostic also carries ready made plots. The precision and recall curve
shows the whole tradeoff instead of a single operating point.

```{r precrec-plot, fig.alt = "Precision and recall plotted against the score threshold for one template."}
diags[[1]]$precrec_plot
```

The density plot separates the score distribution of true positives from that
of false positives. A wide separation means the template discriminates well,
and a heavy overlap means no threshold will work cleanly. Both plots come from
the default run, so their cut line sits to the right of the data, which is the
same result the table above reported in numbers.

```{r density-plot, fig.alt = "Score density of true positives compared with false positives for one template."}
diags[[1]]$plot_dens
```

## Putting the thresholds to work

The diagnostics end with one threshold per template. The last step of the
analysis applies them to the raw detections. `filter_detections()` takes a
threshold table (the template column plus `min_score`, `min_quant` or `top_n`)
and keeps the detections that clear it. Build that table from the relaxed fit
and filter the bundled detections:

```{r apply-thresholds}
thresholds <- data.frame(
  template_name = names(diags_relaxed),
  min_score = sapply(diags_relaxed, `[[`, "score_cut"),
  row.names = NULL
)

df_kept <- filter_detections(df_detecs, thresholds)

c(
  candidates_before = nrow(df_detecs),
  candidates_kept = nrow(df_kept)
)
```

This is where the workflow closes: the kept detections are the validated
template set, filtered at the threshold each template earned in the
diagnostics. They are ready for use on a wider dataset. From here, the next
step is the analysis of that wider dataset: run `template_matching()` on it
with each template's threshold as `min_score`, and use the detections that
clear the score as data for your biological analysis. A small reviewed sample
of the new detections, in `launch_validation_app()`, keeps you sure the
threshold still holds on new recording conditions.

## Templates and spectrograms

The six templates of this example are the Substructure C ROIs: the final
part of each song, one per song in each focal recording. The other focal
ROIs mark the two narrower parts of the same song, Substructure A and
Substructure B. The song lasts about 1.5 seconds and has three parts: two
narrow-band parts between roughly 3.5 and 6 kHz, and a final wider-band part
that extends to about 9 kHz. The final part is Substructure C, so this
example matches on the most striking portion of the song. To match on a
smaller part instead, use the Substructure A or B ROIs as your templates the
same way.

Templates ship as `tuneR` `Wave` objects in `ls_templates`, so the
spectrogram tools run without any file on disk. `extract_spectro()` computes
the spectrogram and `plot_spectro()` draws it.

```{r spectrogram, fig.alt = "Spectrogram of one Basileuterus culicivorus song template."}
data(ls_templates)

spectro <- extract_spectro(ls_templates[[1]])
plot_spectro(spectro)
```

`fast_spectro()` goes from a `Wave` object or a file path straight to the plot
when you do not need the intermediate object.

## Reviewing detections interactively

`launch_validation_app()` opens the counterpart of the segmentation app. It
plays each detection, shows it on the spectrogram and records your verdict,
which is how a manual validation set is built when no ground truth exists yet.

```{r validation-app, eval = FALSE}
launch_validation_app()
```

`detecs_to_rois()` converts validated detections back into ROIs, so a
reviewed run can seed the ground truth of the next one.

## What was created

After you run the workflow on the downloaded project, the project folder
contains:

```
bcu/
├── bcu.qmd                        the walkthrough (already there)
├── bcu.html                       the rendered document
├── recordings/                    the 2 focal recordings (already there)
├── soundscapes/                   the 12 search soundscapes (already there)
├── rois.duckdb                    the 71 ROIs (already there)
├── soundscapes_metadata.duckdb    ← created when metadata were read
├── templates/                     ← created when templates were cut
│   ├── templates.duckdb           the template database: 1 row per template
│   └── Bcu_1_…wav                 the 6 audio cuts
└── detections.duckdb              ← created only if you used output_db
```

| File | Who creates it | What it is for |
|---|---|---|
| `soundscapes_metadata.duckdb` | `fetch_soundscape_metadata()` | duration, sample rate, date/time and path of each recording; a fast re-read instead of scanning the audio again |
| `templates/*.wav` | `export_templates()` | the cut templates the search uses as reference |
| `templates/templates.duckdb` | `export_templates()` | the template database describing the cuts; `fetch_template_metadata()` reads it, directly or through the `templates/` folder |
| `detections.duckdb` | `template_matching(output_db =)` | detections saved to disk, when the session result alone is not enough |

None of these files needs to be kept: if you delete any of them, the commands
above recreate it identically. The only irreplaceable files are the audio
files and `rois.duckdb`, which holds the manual segmentation the templates
depend on.

## Next step

The detections are candidates, not results. `validate_by_overlap()` judges
each one against reference ROIs and `diagnostic_validations()` fits a
threshold to that judgment; both ran above on the bundled tables, and they run
the same way on your own. The workflow is now complete: the templates are
validated and each one carries its threshold, applied to the raw detections
with `filter_detections()` in the previous section. The next step is the wider
dataset: run the matching there with the same thresholds, filter the
resulting detections by score, and use the ones that clear the threshold as
data for your biological analysis. Every function in the workflow also has a
worked example on its help page: the detection engine in `?run_matching`, the
peak picking rules in `?fetch_score_peaks` and the diagnostic definitions in
`?diagnostic_validations`.